Continuation Passing Style for Effect Handlers

### 2020-01-07 docs/solution.tex
Writing article...

Implementation of Continuation Passing Style for Effect Handlers paper \cite{handlers-cps}

@@

    Frank \cite{frank}, Koka \cite{leijen-koka}, Helium \cite{helium}

@@

@@

    \begin{itemize}
        \item Install dependencies: \verb!make install!
        \item Compile: \verb!make build!
        \item Link to PATH: \verb!sudo make link!
        \item Remove artiacts: \verb!make clean!
        \item Run tests: \verb!make tests!
        \item Run code linter: \verb!make lint!
    \end{itemize}

@@

    Test cases are available \href{https://github.com/Tomatosoup97/freak/blob/master/src/Tests.hs}{\underline{here}},
    they include both inline and file-based tests. For more details about
    writing tests, one may refer to HUnit documentation \cite{hunit-docs}. \\

@@

    The Freak implementation is available at \href{https://github.com/Tomatosoup97/freak}{https://github.com/Tomatosoup97/freak}.
    CPS based on \cite{appel-continuations}

    \subsection{Row types}

@@

    Type system as of this day is not implemented, as the focus has been put
    on CPS transformation. Further work is required.

@@

    In this section we present a few examples to show the capabilities of the
    language. The ideas has been based on \cite{eff}, and thus will not be
    described in great details.

@@

    \begin{verbatim}
    handle
        if x == 0 then do Raise ()
                  else return 1 / x
    with {
        Raise p r -> return 42 |
        return x -> return x
    }
    \end{verbatim}

    Where we imagine that x variable has been bound previously. More exemplary
    programs in Freak language can be found \href{https://github.com/Tomatosoup97/freak/tree/master/src/programs}{\underline{here}}.

@@

    \subsection{Abstract machine}

    Implement abstract machine described in \cite{liberating-effects}

    Proposed Helium \cite{binders-labels}, or Eff \cite{eff}

@@

@@

    As in \cite{leijen-koka}

@@

    X \cite{shallow-handlers}

>>>>

In this article we present experimental programming language Freak, which is an
implementation of Continuation Passing Style for Effect Handlers paper \cite{handlers-cps},
with addition of a few basic constructs. We start by presenting the related work,
then discuss syntax and operational semantics. Basic usage guide for playing with
the language is provided, as well as implementation details and examples. We conclude
by stating what are the possible augmentations, that are intended to be made in the future.

@@

    Despite from Links language \cite{handlers-cps}, on which the implementation
    is based, there are currently many other alternatives available. One may take
    a look at Frank \cite{frank}, which provides a support for multihandlers,
    Koka \cite{leijen-koka}, Helium \cite{helium} or Eff \cite{eff}. Except from
    separate languages, many libraries arose for existing ones like Haskell,
    Idris, Scala or Multicore OCaml. \\

    As can be seen in the J. Yallop repository \cite{effects-bibliography},
    algebraic effects and handlers are now trending branch of computer science.

@@

    The source language's dynamics has been described
    extensively by providing small-step operational semantics,
    continuation passing style transformation \cite{handlers-cps} as well
    as abstract machine \cite{liberating-effects}, which was proved to coincide
    with CPS translation. That being said, Freak introduces new basic
    constructs to the language, for which we shall define the semantics.

@@

    \begin{itemize}
        \item Install dependencies: \verb!make install!
        \item Compile: \verb!make build!
        \item Link to PATH: \verb!sudo make link!
        \item Remove artiacts: \verb!make clean!
    \end{itemize}

@@

    Test cases are available \href{https://github.com/Tomatosoup97/freak/blob/master/src/Tests.hs}{\underline{here}},
    they include both inline and file-based tests. For more details about
    writing tests, one may refer to \textit{HUnit documentation} \cite{hunit-docs}. \\

@@

    The Freak implementation is available at \href{https://github.com/Tomatosoup97/freak}{https://github.com/Tomatosoup97/freak}.
    While the paper provided a good overview on the language and the translation,
    the lower-level details were ommited. Therefore, the actual CPS implementation
    is based on A. Appel \cite{appel-continuations}, which provides a translation
    for a simplified ML calculus. \\

    The core data structure into which the source program is transformed,
    the \verb!Continuation Computation!, is defined as follows:

    \begin{verbatim}
data ContComp
    = CPSApp CValue [CValue]
    | CPSFix Var [Var] ContComp ContComp
    | CPSBinOp BinaryOp CValue CValue Var ContComp
    | CPSValue CValue
    | CPSLet Var CValue ContComp
    | CPSSplit Label Var Var CValue ContComp
    | CPSCase CValue Label Var ContComp Var ContComp
    | CPSIf CValue ContComp ContComp
    | CPSAbsurd CValue
    \end{verbatim}

    The translation is based on the curried first-order translation. For
    improving the performance of the evaluation, uncurried higher-order translation
    should be adopted, so that administrative redexes are contracted and
    proper tail-recursion is obtained. The final answer is represented as a \verb!DValue!,
    where the meaning of the coproduct is as one would expect:

    \begin{verbatim}
type Label = String

data DValue
    = DNum Integer
    | DLambda Env FuncRecord
    | DUnit
    | DPair DValue DValue
    | DLabel Label
    \end{verbatim}

@@

    Type system as of this day is not implemented, as the focus has been put
    on CPS transformation. Further work is required here, especially considering
    the fact that a huge advantage of algebraic effects is that they are explicitly
    defined in the type of a computation.

    \subsection{Source code structure}

    The source code is divided into a number of modules, each with a different
    responsibility.

    \begin{verbatim}
    AST.hs          - AST data structures
    CPS.hs          - CPS translation
    Eval.hs         - Evaluation of the target calculus
    Freak.hs        - API for the language
    Main.hs         - Main module running evaluator on given filename
    Parser.hs       - Parser and lexer
    Tests.hs        - Tests module
    Typechecker.hs  - -- suspended --
    Types.hs        - Common types definition
    programs/       - Exemplary programs used in tests
    \end{verbatim}

@@

    In this section we present a few examples to show the capabilities of the
    language. The ideas has been based on \cite{programming-in-eff}, and thus will not be
    described in great details. More exemplary programs in Freak language can
    be found \href{https://github.com/Tomatosoup97/freak/tree/master/src/programs}{\underline{here}}.

@@

    \begin{verbatim}
    handle
        if x == 0 then do ZeroDivisionError ()
                  else return 1/x
    with {
        ZeroDivisionError p r -> return 42 |
        return x -> return x
    }
    \end{verbatim}
    Where we imagine that x variable has been bound previously.

@@

    The Freak language is experimental and a lot of possible enhancements
    could be adopted. Here we provide a list of proposed directions in which
    further work could be made.

    \subsection{Alternative evaluation}

    The Links language also provides small-step operational semantics and
    an abstract machine \cite{liberating-effects}. Implementing another way
    of evaluation could serve as a way to empirically assert correctness,
    as opposed to formally.

@@

    The Freak language is limited to a single instance of an effect. We would
    need to support cases where many instances of the algebraic effects, with
    the same handlers code, could be instantiated. The current state of the
    art introduces a concept of resources and instances, as in Eff \cite{programming-in-eff},
    or instance variables, as in Helium \cite{binders-labels}.

@@

    Other languages, like Koka \cite{leijen-koka}, or even core of the Links, are
    performing selective CPS translation, which reduces the overhead on code
    that does not perform algebraic effects. Our current translation is fully
    embeded in the CPS.

@@

    Shallow and deep handlers while being able to simluate each other up to
    administrative reductions, have a very different meaning from a theoretical
    point of view. Implementing them as defined by Lindley et al. \cite{shallow-handlers} could
    be another way of enhancing Freak.
